Refunctionalization of Abstract Abstract Machines (Functional Pearl)

### 2018-06-05 paper/paper.tex
refactor function tick

\subsubsection{Allocating Addresses}
Up to this point, we have not described allocating addresses in stores,
nor handling the time stamp @Time@.
In abstract interpretation, however, these are key ingredients to achieve
analyses with different sensitivities, as well as to perform a finite
state space analysis\cite{Gilray:2016:ACP:2951913.2951936}.
To effectively approximate the runtime behavior, we introduce a finite
program contour @time@ that encodes the program execution history.
We use a list of execution contexts (expressions) to represent this, and
as we will see in Section~\ref{kcfainst}, by applying different @tick@
functions on the timestamp, we are able to obtain a family of analyses.

@@

\begin{lstlisting}
def step(s: State): List[State] = {
  val new_time = tick(s)
  s match {
    case State(Let(x, App(f, ae), e), env, bstore, kstore, kaddr, time) =>
      val closures = atomicEval(f, env, bstore).toList
      for (Clos(Lam(v, body), env_c) <- closures) yield {
        val baddr = allocBind(v, new_time)
        val new_env = env_c + (v -> baddr)
        val new_bstore = bstore.update(baddr, atomicEval(ae, env, bstore))
        val new_kaddr = allocKont(body)
        val new_kstore = kstore.update(new_kaddr, Cont(Frame(x, e, env), kaddr))
        State(body, new_env, new_bstore, new_kstore, new_kaddr, new_time)
      }
    case State(ae, env, bstore, kstore, kaddr, time) if isAtomic(ae) =>
      val conts = kstore(kaddr).toList
      for (Cont(Frame(x, e, env_f), f_kaddr) <- conts) yield {
        val baddr = allocBind(x, new_time)
        val new_env = env_f + (x -> baddr)
        val new_store = bstore.update(baddr, atomicEval(ae, env, bstore))
        State(e, new_env, new_store, kstore, f_kaddr, new_time)
      }
  }
}
\end{lstlisting}

@@

Before transferring to successor states, we must use the @tick@ function to refresh the 
timestamp, and then use this new timestamp for successors when allocating addresses.
The @tick@ function returns the $k$ front-most expressions given the current state and its
time history.

\begin{lstlisting}
def k: Int = 0
def tick(s: State): Time = (s.e :: s.time).take(k)
\end{lstlisting}

@@

\begin{lstlisting}
def step(s: State): List[State] = {
  val new_time = tick(s)
  s match {
    case State(Let(x, App(f, ae), e), env, bstore, konts, time) if isAtomic(ae) =>
      for (Clos(Lam(v, body), env_c) <- atomicEval(f, env, bstore).toList) yield {
        val frame = Frame(x, e, env)
        val baddr = allocBind(v, new_time)
        val new_env = env_c + (v -> baddr)
        val new_store = bstore.update(baddr, atomicEval(ae, env, bstore))
        State(body, new_env, new_store, frame::konts, new_time)
      }
    case State(ae, env, bstore, konts, time) if isAtomic(ae) =>
      konts match {
        case Nil => List()
        case Frame(x, e, env_f)::konts =>
          val baddr = allocBind(x, new_time)
          val new_env = env_f + (x -> baddr)
          val new_store = bstore.update(baddr, atomicEval(ae, env, bstore))
          List(State(e, new_env, new_store, konts, new_time))
      }
  }
}
\end{lstlisting}

@@

\begin{lstlisting}
def drive_step(nds: NDState, seen: Set[State]): Set[State] = {
  nds match {
    case NDState(ae, _, _, Nil, _, Nil) if isAtomic(ae) => seen
    case nds =>
      val s = nds.toState
      val new_seen = if (seen.contains(s)) seen else seen+s
      val newTime = tick(nds)
      val new_ndstate = nds match {
        case NDState(Let(x, App(f, ae), e), env, bstore, konts, time, ndkonts) =>
          ......
        case NDState(ae, env, bstore, konts, time, ndkonts) if isAtomic(ae) =>
          ......
      }
      drive_step(new_ndstate, new_seen)
  }
}
\end{lstlisting}

@@

\begin{lstlisting}
def drive_step(nds: NDState, seen: Set[State]): Set[State] = {
  nds match {
    case NDState(ae, _, _, Nil, _, Nil) if isAtomic(ae) => seen
    case nds =>
      val s = nds.toState
      val new_seen = if (seen.contains(s)) seen else seen + s
      val new_time = tick(nds)
      nds match {
        case NDState(Let(x, App(f, ae), e), env, bstore, konts, time, ndk) =>
          ......
          drive_step(NDState(body, new_env, new_store, new_frames, new_time, new_ndk), new_seen)
        case NDState(ae, env, bstore, konts, time, ndk) if isAtomic(ae) =>
          continue(nds, new_seen)
      }
  }
}
\end{lstlisting}

@@

\begin{lstlisting}
def continue(nds: NDState, seen: Set[State]): Set[State] = {
  val NDState(ae, env, bstore, konts, time, ndk) = nds
  val new_time = tick(nds)
  konts match {
    case Nil => ndcontinue(nds, seen)
    case Frame(x, e, f_env)::konts =>
      val baddr = allocBind(x, new_time)
      val new_env = f_env + (x -> baddr)
      val new_store = bstore.update(baddr, atomicEval(ae, env, bstore))
      drive_step(NDState(e, new_env, new_store, konts, new_time, ndk), seen)
  }
}
\end{lstlisting}

@@

To properly represent final values, we introduce the case class @VS@ which
contains a set of storable values, a timestamp, and a store.
\begin{lstlisting}
case class VS(vals: Set[Storable], time: Time, store: BStore)
\end{lstlisting}
An instance of @VS@ represents the computational result of one path in
the nondeterministic evaluation.
The reason that we include a store is that there might be a case in which two paths
have the same values but different accumulated side effects (e.g., memory allocations).
Since the whole computation is nondeterministic, the type of the final result
values is a collection @Set[VS]@. For convenience, we define it as a type @Ans@:
\begin{lstlisting}
type Ans = Set[VS]
\end{lstlisting}

As previously mentioned, the components of states are lifted as arguments to
@aeval@. Therefore, the pattern matching on the state becomes just a match on
expression @e@.
\begin{lstlisting}
type Cont = Ans => Ans
def aeval(e: Expr, env: Env, store: BStore, time: Time, continue: Cont): Ans = {
  val new_time = (e::time).take(k)
  e match {
    case Let(x, App(f, ae), e) =>
      ......
    case ae if isAtomic(ae) =>
      ......
  }
}
\end{lstlisting}

@@

\textbf{Configuration.}
We do not have the concept of an explicit stack or state,
so to represent the cache we introduce a configuration @Config@ as
a state-like definition which borrows the components from state, sans continuations.
\begin{lstlisting}
case class Config(e: Expr, env: Env, store: BStore, time: Time)
\end{lstlisting}

@@

\begin{lstlisting}
def aeval(e: Expr, env: Env, store: BStore, time: Time, cache: Cache, continue: Cont): Ans = {
  val config = Config(e, env, store, time)
  /* lookups the out cache, uses the values if contains this config */
  if (cache.outContains(config)) cont(Ans(cache.outGet(config), cache))
  else {
    val new_cache = cache.outUpdate(config, cache.inGet(config))
    val new_time = (e::time).take(k)
    e match {
      case Let(x, App(f, ae), e) =>
        val closures = atomicEval(f, env, store).asInstanceOf[Set[Clos]]
        nd[Clos, Ans](closures, Ans(Set[VS](), new_cache), { 
          case (Clos(Lam(v, body), c_env), clsans, clsnd) =>
            ...... /* sets vairables, e.g., new_cenv and new_store */
            aeval(body, new_cenv, new_store, new_time, clsans.cache, { case Ans(bdvss, bdcache) =>
              nd[VS, Ans](bdvss, Ans(Set[VS](), bdcache), { 
                case (VS(vals, time, vssstore), bdans, bdnd) =>
                  ...... /* sets variables, e.g., new_env and new_store */
                  aeval(e, new_env, new_store, time, bdans.cache, { 
                    /* updates the out cache after evaluated to the end of one path  */
                    case Ans(evss, ecache) => bdnd(bdans ++ Ans(evss, ecache.outUpdate(config, evss)))
                  })
              }, { case eans => clsnd(eans ++ clsans) })
            })
        }, 
        cont)

@@

\begin{lstlisting}
def aeval(e: Expr, env: Env, store: BStore, time: Time, cache: Cache): Ans @cps[Ans] = {
  val config = Config(e, env, store, time)
  if (cache.outContains(config)) Ans(cache.outGet(config), cache)
  else {
    val new_time = (e::time).take(k)
    val new_cache = cache.outUpdate(config, cache.inGet(config))
    e match {
      case Let(x, App(f, ae), e) =>
        val closures = atomicEval(f, env, store).asInstanceOf[Set[Clos]]
        val (Clos(Lam(v, body), c_env), clscache) = ndcps[Clos](closures, Ans(Set[VS](), new_cache))
        val vbaddr = allocBind(v, new_time)
        val new_cenv = c_env + (v -> vbaddr)
        val new_store = store.update(vbaddr, atomicEval(ae, env, store))
        /* evaluates the function application App(f, ae) */
        val Ans(bdvss, bdcache) = aeval(body, new_cenv, new_store, new_time, clscache)
        val (VS(vals, time, vsstore), vscache) = ndcps[VS](bdvss, Ans(Set[VS](), bdcache))
        val baddr = allocBind(x, time)
        val new_env = env + (x -> baddr)
        val new_vsstore = vsstore.update(baddr, vals)
        /* evaluates the body expression from the let */
        val Ans(evss, ecache) = aeval(e, new_env, new_vsstore, time, vscache)
        Ans(evss, ecache.outUpdate(config, evss))

>>>>

\subsubsection{Allocating Addresses}
Up to this point, we have not described allocating addresses in stores,
nor handling the time stamp @Time@.
In abstract interpretation, however, these are key ingredients to achieve
analyses with different sensitivities, as well as to perform a finite
state space analysis\cite{Gilray:2016:ACP:2951913.2951936}.
To effectively approximate the runtime behavior, we introduce a finite
program contour @time@ that encodes the program execution history.
The function @tick@ is used to refresh the "time" and get a "new time".
We use a list of execution contexts (expressions) to represent the finite 
program contour, and as we will see in Section~\ref{kcfainst}, by applying 
different @tick@ functions on the timestamp, we are able to obtain a 
family of analyses.

@@

\begin{lstlisting}
def step(s: State): List[State] = {
  val new_time = s.tick
  s match {
    case State(Let(x, App(f, ae), e), env, bstore, kstore, kaddr, time) =>
      val closures = atomicEval(f, env, bstore).toList
      for (Clos(Lam(v, body), env_c) <- closures) yield {
        val baddr = allocBind(v, new_time)
        val new_env = env_c + (v -> baddr)
        val new_bstore = bstore.update(baddr, atomicEval(ae, env, bstore))
        val new_kaddr = allocKont(body)
        val new_kstore = kstore.update(new_kaddr, Cont(Frame(x, e, env), kaddr))
        State(body, new_env, new_bstore, new_kstore, new_kaddr, new_time)
      }
    case State(ae, env, bstore, kstore, kaddr, time) if isAtomic(ae) =>
      val conts = kstore(kaddr).toList
      for (Cont(Frame(x, e, env_f), f_kaddr) <- conts) yield {
        val baddr = allocBind(x, new_time)
        val new_env = env_f + (x -> baddr)
        val new_store = bstore.update(baddr, atomicEval(ae, env, bstore))
        State(e, new_env, new_store, kstore, f_kaddr, new_time)
      }
  }
}
\end{lstlisting}

@@

Before transferring to successor states, we must use the @tick@ function to refresh the 
timestamp, and then use this new timestamp for successors when allocating addresses.
The @tick@ function returns the $k$ front-most expressions given the current state and its
time history. We implement @tick@ as a public method of case class @State@.

\begin{lstlisting}
def k: Int = 0
case class State(e: Expr, env: Env, bstore: BStore, kstore: KStore, kaddr: KAddr, time: Time) {
  def tick: Time = (e :: time).take(k)
}
\end{lstlisting}

@@

\begin{lstlisting}
def step(s: State): List[State] = {
  val new_time = s.tick
  s match {
    case State(Let(x, App(f, ae), e), env, bstore, konts, time) if isAtomic(ae) =>
      for (Clos(Lam(v, body), env_c) <- atomicEval(f, env, bstore).toList) yield {
        val frame = Frame(x, e, env)
        val baddr = allocBind(v, new_time)
        val new_env = env_c + (v -> baddr)
        val new_store = bstore.update(baddr, atomicEval(ae, env, bstore))
        State(body, new_env, new_store, frame::konts, new_time)
      }
    case State(ae, env, bstore, konts, time) if isAtomic(ae) =>
      konts match {
        case Nil => List()
        case Frame(x, e, env_f)::konts =>
          val baddr = allocBind(x, new_time)
          val new_env = env_f + (x -> baddr)
          val new_store = bstore.update(baddr, atomicEval(ae, env, bstore))
          List(State(e, new_env, new_store, konts, new_time))
      }
  }
}
\end{lstlisting}

@@

\begin{lstlisting}
def drive_step(nds: NDState, seen: Set[State]): Set[State] = {
  nds match {
    case NDState(ae, _, _, Nil, _, Nil) if isAtomic(ae) => seen
    case nds =>
      val s = nds.toState
      val new_seen = if (seen.contains(s)) seen else seen+s
      val newTime = nds.tick
      val new_ndstate = nds match {
        case NDState(Let(x, App(f, ae), e), env, bstore, konts, time, ndkonts) =>
          ......
        case NDState(ae, env, bstore, konts, time, ndkonts) if isAtomic(ae) =>
          ......
      }
      drive_step(new_ndstate, new_seen)
  }
}
\end{lstlisting}

@@

\begin{lstlisting}
def drive_step(nds: NDState, seen: Set[State]): Set[State] = {
  nds match {
    case NDState(ae, _, _, Nil, _, Nil) if isAtomic(ae) => seen
    case nds =>
      val s = nds.toState
      val new_seen = if (seen.contains(s)) seen else seen + s
      val new_time = nds.tick
      nds match {
        case NDState(Let(x, App(f, ae), e), env, bstore, konts, time, ndk) =>
          ......
          drive_step(NDState(body, new_env, new_store, new_frames, new_time, new_ndk), new_seen)
        case NDState(ae, env, bstore, konts, time, ndk) if isAtomic(ae) =>
          continue(nds, new_seen)
      }
  }
}
\end{lstlisting}

@@

\begin{lstlisting}
def continue(nds: NDState, seen: Set[State]): Set[State] = {
  val NDState(ae, env, bstore, konts, time, ndk) = nds
  val new_time = nds.tick
  konts match {
    case Nil => ndcontinue(nds, seen)
    case Frame(x, e, f_env)::konts =>
      val baddr = allocBind(x, new_time)
      val new_env = f_env + (x -> baddr)
      val new_store = bstore.update(baddr, atomicEval(ae, env, bstore))
      drive_step(NDState(e, new_env, new_store, konts, new_time, ndk), seen)
  }
}
\end{lstlisting}

@@

To properly represent final values, we introduce the case class @VS@ which
contains a set of storable values, a timestamp, and a store.

\begin{lstlisting}
case class VS(vals: Set[Storable], time: Time, store: BStore)
\end{lstlisting}

An instance of @VS@ represents the computational result of one path in
the nondeterministic evaluation.
The reason that we include a store is that there might be a case in which two paths
have the same values but different accumulated side effects (e.g., memory allocations).
Since the whole computation is nondeterministic, the type of the final result
values is a collection @Set[VS]@. For convenience, we define it as a type @Ans@:

\begin{lstlisting}
type Ans = Set[VS]
\end{lstlisting}

As previously mentioned, the components of states are lifted as arguments to
@aeval@. Therefore, the pattern matching on the state becomes just a match on
expression @e@. The shape of function @aeval@ looks like:

\begin{lstlisting}
type Cont = Ans => Ans
def aeval(e: Expr, env: Env, store: BStore, time: Time, continue: Cont): Ans = {
  val new_time = ......
  e match {
    case Let(x, App(f, ae), e) => ......
    case ae if isAtomic(ae) => ......
  }
}
\end{lstlisting}

@@

\textbf{Configuration.}
We do not have the concept of an explicit stack or state,
so to represent the cache we introduce a configuration @Config@ as
a state-like definition which borrows the components from state, sans continuations.

\begin{lstlisting}
case class Config(e: Expr, env: Env, store: BStore, time: Time)
\end{lstlisting}

@@

\begin{lstlisting}
def aeval(e: Expr, env: Env, store: BStore, time: Time, cache: Cache, continue: Cont): Ans = {
  val config = Config(e, env, store, time)
  /* lookups the out cache, uses the values if contains this config */
  if (cache.outContains(config)) cont(Ans(cache.outGet(config), cache))
  else {
    val new_cache = cache.outUpdate(config, cache.inGet(config))
    val new_time = config.tick 
    e match {
      case Let(x, App(f, ae), e) =>
        val closures = atomicEval(f, env, store).asInstanceOf[Set[Clos]]
        nd[Clos, Ans](closures, Ans(Set[VS](), new_cache), { 
          case (Clos(Lam(v, body), c_env), clsans, clsnd) =>
            ...... /* sets vairables, e.g., new_cenv and new_store */
            aeval(body, new_cenv, new_store, new_time, clsans.cache, { case Ans(bdvss, bdcache) =>
              nd[VS, Ans](bdvss, Ans(Set[VS](), bdcache), { 
                case (VS(vals, time, vssstore), bdans, bdnd) =>
                  ...... /* sets variables, e.g., new_env and new_store */
                  aeval(e, new_env, new_store, time, bdans.cache, { 
                    /* updates the out cache after evaluated to the end of one path  */
                    case Ans(evss, ecache) => bdnd(bdans ++ Ans(evss, ecache.outUpdate(config, evss)))
                  })
              }, { case eans => clsnd(eans ++ clsans) })
            })
        }, 
        cont)

@@

\begin{lstlisting}
def aeval(e: Expr, env: Env, store: BStore, time: Time, cache: Cache): Ans @cps[Ans] = {
  val config = Config(e, env, store, time)
  if (cache.outContains(config)) Ans(cache.outGet(config), cache)
  else {
    val new_time = config.tick
    val new_cache = cache.outUpdate(config, cache.inGet(config))
    e match {
      case Let(x, App(f, ae), e) =>
        val closures = atomicEval(f, env, store).asInstanceOf[Set[Clos]]
        val (Clos(Lam(v, body), c_env), clscache) = ndcps[Clos](closures, Ans(Set[VS](), new_cache))
        val vbaddr = allocBind(v, new_time)
        val new_cenv = c_env + (v -> vbaddr)
        val new_store = store.update(vbaddr, atomicEval(ae, env, store))
        /* evaluates the function application App(f, ae) */
        val Ans(bdvss, bdcache) = aeval(body, new_cenv, new_store, new_time, clscache)
        val (VS(vals, time, vsstore), vscache) = ndcps[VS](bdvss, Ans(Set[VS](), bdcache))
        val baddr = allocBind(x, time)
        val new_env = env + (x -> baddr)
        val new_vsstore = vsstore.update(baddr, vals)
        /* evaluates the body expression from the let */
        val Ans(evss, ecache) = aeval(e, new_env, new_vsstore, time, vscache)
        Ans(evss, ecache.outUpdate(config, evss))
